An enhancer code compendium of early human neural and neural crest development

### 2025-05-12 manuscript.tex
KRT refs.

@@

\subsubsection*{\textit{de novo} motif discovery}
For all organoid and embryo topics using respectively $DeepNeuralTube_O$ and $DeepNeuralTube_E$ attribution scores
were calculated using the CREsted\cite{kempynckCREstedModelingGenomic2025} (v1.3.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026617/resolver}{RRID:SCR\_026617}) function integrated\_grad, on 1,000 regions with the highest prediction score for that topic according to the model. Next, patterns/motifs were discovered \textit{de novo} using a custom implementation of tfmodisco-lite (v2.0.0; RRID:SCR\_026709). First seqlets were called using the tfmodisco-lite function extract\_seqlets using the attribution scores as input and using parameters 
sliding\_window\_size = 15, 
flank = 5, 
suppress = 12, 
target\_fdr = 0.2, 
min\_passing\_windows\_frac = 0.03, 
max\_passing\_windows\_frac = 0.3 and 
weak\_threshold\_for\_counting\_sig = 0.8. 
Next, the attribution score of each seqlet was extracted based on the seqlet coordinates and negative and positive seqlets were defined based on the sum of the attribution score without the flank filtering out seqlets for which the abolute value of the sum of the attribution score was lower than or equal to X. 
Finally, separately for positive and negative seqlets 1,000 seqlets were sampled and used in the tfmodisco-lite function seqlet\_to\_patterns using parameters 
min\_overlap\_while\_sliding = 0.7,
nearest\_neighbors\_to\_compute = 500,
affmat\_correlation\_threshold = 0.15,
tsne\_perplexity = 10,
n\_leiden\_iterations = -1,
n\_leiden\_runs = 50,
frac\_support\_to\_trim\_to = 0.2,
min\_num\_to\_trim\_to = 30,
trim\_to\_window\_size = 20,
initial\_flank\_to\_add = 5,
prob\_and\_pertrack\_sim\_merge\_thresholds = [(0.8,0.8), (0.5, 0.85), (0.2, 0.9)],
prob\_and\_pertrack\_sim\_dealbreaker\_thresholds = [(0.4, 0.75), (0.2,0.8), (0.1, 0.85), (0.0,0.9)],
subcluster\_perplexity = 50,
merging\_max\_seqlets\_subsample = 300,
final\_min\_cluster\_size = 20,
min\_ic\_in\_window = 0.6,
min\_ic\_windowsize = 6 and 
ppm\_pseudocount= 0.001.
Next, similar patterns were clustered together by first calculating a pairwise pattern-to-pattern similarity metric
using TomTom\cite{guptaQuantifyingSimilarityMotifs2007} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}). The Pearson correlation of one minus the E-value (product of the TomTom p-value and the number of patterns), calculated using the corrcoef function from NumPy\cite{harrisArrayProgrammingNumPy2020} (v2.0.1; \href{https://rrid.site/data/record/nlx_144509-1/SCR_008633/resolver}{RRID:SCR\_008633}), was used for hierarchical clustering using the hierarchy.linkage (using average as method) and fcluster (using maxclust as criterion and a manual defined number of clusters) functions from Scipy\cite{virtanenSciPy10Fundamental2020} (v1.15.2; \href{https://rrid.site/data/record/nlx_144509-1/SCR_008058/resolver}{RRID:SCR\_008058}).

@@

Enhancer candidates were ordered from Twist Bioscience (\href{https://rrid.site/data/record/nlx_144509-1/SCR_025817/resolver}{RRID:SCR\_025817}) with 5\lq 5\lq-GGTACCGAGCTCGAG-3\lq and 3\lq 5\lq-GGGCTCGAGATCTGC-3\lq adaptors.
Plasmid pTK BsmBI LacZ Citrine Nanotag 24 (\href{https://rrid.site/data/record/nif-0000-11872-1/Addgene_130522/resolver}{RRID:Addgene\_130522}) was linarized using forward primer 5\lq-ggAACTCGAGCTCGGTACCT-3\lq and reverse primer 5\lq-GGGCTCGAGATCTGCGATC-3\lq using PCR. A PCR mixture of KAPA HiFi hot start ready mix (Roche, 07958935001 \SI{50}{\percent} V/V), \SI{0.8}{\micro\gram\per\micro\litre} plasmid and \SI{0.3}{\micro\molar} of forward and reverse primer and 
PCR program: \SI{95}{\degreeCelsius} for 3 minutes, 15 cycles of: \SI{98}{\degreeCelsius} for 20 seconds,
\SI{58}{\degreeCelsius} for 15 seconds and \SI{72}{\degreeCelsius} for 2.5 minutes, followed by \SI{72}{\degreeCelsius}
for 5 minutes was used.
Linarized plasmids were purified using QIAquick PCR purification kit (Qiagen, 28104), followed by DpnI digestion for 1
hour at \SI{37}{\degreeCelsius} and gel extraction of 4,5 kb band using QIAquick Gel Extraction Kit (Qiagen, 28704).
Enhancer candiates were cloned into the target vector using an NEBuilder (New England Biolabs, E5520S) reaction with a linarized plasmid to insert ratio of 1:2 (45 minutes at \SI{50}{\degreeCelsius}). Next, \SI{2.5}{\micro\litre} of the rection mixture was transformed into \SI{25}{\micro\litre} Stellar Competent Cells (Takara Bio, 636763). Cells were thawed 15 minutes on ice, plasmid + cell mixture was kept for an additional 30 minutes on ice followed by 45 second heat shot at \SI{42}{\degreeCelsius} and 5 minutes on ice. 
Cells were incubutated for 1 hour at \SI{37}{\degreeCelsius}, single colonies were then grown on Carbencillin plates
over night.
Finally, plasmids were purified using NucleoBond Xtra Maxi Plus Kit (Macherey-Nagel, 740416.50) and eluded at a target
concentration of \SI{4}{\micro\gram\per\micro\litre}.

\paragraph*{Injection control cloning}
Plasmid pCI H2B-RFP (\href{https://rrid.site/data/record/nif-0000-11872-1/Addgene_92398/resolver}{RRID:Addgene\_92398})
was ordered from Addgene (\href{https://rrid.site/data/record/nlx_144509-1/SCR_002037/resolver}{RRID:SCR\_002037}),
transformed in Stellar Competent Cells (Takara Bio, 636763). Cells were thawed 15 minutes on ice, plasmid + cell mixture was kept for an additional 30 minutes on ice followed by 45 second heat shot at \SI{42}{\degreeCelsius} and 5 minutes on ice. 
Cells were incubutated for 1 hour at \SI{37}{\degreeCelsius}, single colonies were then grown on Carbencillin plates
over night.
Finally, plasmids were purified using NucleoBond Xtra Maxi Plus Kit (Macherey-Nagel, 740416.50) and eluded at a target
concentration of \SI{4}{\micro\gram\per\micro\litre}.

@@

\subsubsection*{General and specific FOXA2 target regions}
FOXA2 ChIP-seq peaks in HepG2 (\href{https://www.encodeproject.org/files/ENCFF466FCB/}{ENCFF466FCB}) and A549 \href{https://www.encodeproject.org/files/ENCFF686MSH/}{ENCFF686MSH} were downloaded from ENCODE (\href{https://rrid.site/data/record/nlx_144509-1/SCR_006793/resolver}{RRID:SCR\_006793}). 
For each pattern (converted to a PWM) of all neural progenitor pattern cluster motif hits were calculated on the top
1,000 regions of all neural progenitor topics using FIMO\cite{grantFIMOScanningOccurrences2011} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}) using default parameters.
For each sequence the maximum hit, based on the average contribution score of that hit for the topic from which the
region originated, was retained.
Attribution scores were normalized by dividing over the sum of attribution scores over all patterns per sequence.
FOX hits were defined as hits with a value greater than 0.0008 in organoid and 0.0004 in embryo.
To define general hits, the FOX hits were intersected with FOXA2 ChIP-seq peaks in HepG2 and A549 using bedtools (v2.31;
\href{https://rrid.site/data/record/nlx_144509-1/SCR_006646/resolver?q=bedtools&l=bedtools&i=rrid:scr_006646}{RRID:SCR\_006646}) intersect using option -F 0.4 and peaks intersecting with either of the two ChIP-seq peak sets were defined as general.

>>>>

%TODO: add reference to https://www.sciencedirect.com/science/article/pii/S2213671114003580?via%3Dihub and put number in
%KRT for Verfaillie et al. and pierre

@@

\subsubsection*{\textit{de novo} motif discovery}
For all organoid and embryo topics using respectively $DeepNeuralTube_O$ and $DeepNeuralTube_E$ attribution scores
were calculated using the CREsted\cite{kempynckCREstedModelingGenomic2025} (v1.3.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026617/resolver}{RRID:SCR\_026617}) function integrated\_grad, on 1,000 regions with the highest prediction score for that topic according to the model. Next, patterns/motifs were discovered \textit{de novo} using a custom implementation of tfmodisco-lite (v2.0.0; RRID:SCR\_026709)\cite{schreiberTfmodiscolite2025}. First seqlets were called using the tfmodisco-lite function extract\_seqlets using the attribution scores as input and using parameters 
sliding\_window\_size = 15, 
flank = 5, 
suppress = 12, 
target\_fdr = 0.2, 
min\_passing\_windows\_frac = 0.03, 
max\_passing\_windows\_frac = 0.3 and 
weak\_threshold\_for\_counting\_sig = 0.8. 
Next, the attribution score of each seqlet was extracted based on the seqlet coordinates and negative and positive seqlets were defined based on the sum of the attribution score without the flank filtering out seqlets for which the abolute value of the sum of the attribution score was lower than or equal to X. 
Finally, separately for positive and negative seqlets 1,000 seqlets were sampled and used in the tfmodisco-lite function seqlet\_to\_patterns using parameters 
min\_overlap\_while\_sliding = 0.7,
nearest\_neighbors\_to\_compute = 500,
affmat\_correlation\_threshold = 0.15,
tsne\_perplexity = 10,
n\_leiden\_iterations = -1,
n\_leiden\_runs = 50,
frac\_support\_to\_trim\_to = 0.2,
min\_num\_to\_trim\_to = 30,
trim\_to\_window\_size = 20,
initial\_flank\_to\_add = 5,
prob\_and\_pertrack\_sim\_merge\_thresholds = [(0.8,0.8), (0.5, 0.85), (0.2, 0.9)],
prob\_and\_pertrack\_sim\_dealbreaker\_thresholds = [(0.4, 0.75), (0.2,0.8), (0.1, 0.85), (0.0,0.9)],
subcluster\_perplexity = 50,
merging\_max\_seqlets\_subsample = 300,
final\_min\_cluster\_size = 20,
min\_ic\_in\_window = 0.6,
min\_ic\_windowsize = 6 and 
ppm\_pseudocount= 0.001.
Next, similar patterns were clustered together by first calculating a pairwise pattern-to-pattern similarity metric
using TomTom\cite{guptaQuantifyingSimilarityMotifs2007} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}). The Pearson correlation of one minus the E-value (product of the TomTom p-value and the number of patterns), calculated using the corrcoef function from NumPy\cite{harrisArrayProgrammingNumPy2020} (v2.0.1; \href{https://rrid.site/data/record/nlx_144509-1/SCR_008633/resolver}{RRID:SCR\_008633}), was used for hierarchical clustering using the hierarchy.linkage (using average as method) and fcluster (using maxclust as criterion and a manual defined number of clusters) functions from Scipy\cite{virtanenSciPy10Fundamental2020} (v1.15.2; \href{https://rrid.site/data/record/nlx_144509-1/SCR_008058/resolver}{RRID:SCR\_008058}).

@@

Enhancer candidates were ordered from Twist Bioscience (\href{https://rrid.site/data/record/nlx_144509-1/SCR_025817/resolver}{RRID:SCR\_025817}) with 5\lq 5\lq-GGTACCGAGCTCGAG-3\lq and 3\lq 5\lq-GGGCTCGAGATCTGC-3\lq adaptors.
Plasmid pTK BsmBI LacZ Citrine Nanotag 24 (\href{https://rrid.site/data/record/nif-0000-11872-1/Addgene_130522/resolver}{RRID:Addgene\_130522})\cite{williamsReconstructionGlobalNeural2019} was linarized using forward primer 5\lq-ggAACTCGAGCTCGGTACCT-3\lq and reverse primer 5\lq-GGGCTCGAGATCTGCGATC-3\lq using PCR. A PCR mixture of KAPA HiFi hot start ready mix (Roche, 07958935001 \SI{50}{\percent} V/V), \SI{0.8}{\micro\gram\per\micro\litre} plasmid and \SI{0.3}{\micro\molar} of forward and reverse primer and 
PCR program: \SI{95}{\degreeCelsius} for 3 minutes, 15 cycles of: \SI{98}{\degreeCelsius} for 20 seconds,
\SI{58}{\degreeCelsius} for 15 seconds and \SI{72}{\degreeCelsius} for 2.5 minutes, followed by \SI{72}{\degreeCelsius}
for 5 minutes was used.
Linarized plasmids were purified using QIAquick PCR purification kit (Qiagen, 28104), followed by DpnI digestion for 1
hour at \SI{37}{\degreeCelsius} and gel extraction of 4,5 kb band using QIAquick Gel Extraction Kit (Qiagen, 28704).
Enhancer candiates were cloned into the target vector using an NEBuilder (New England Biolabs, E5520S) reaction with a linarized plasmid to insert ratio of 1:2 (45 minutes at \SI{50}{\degreeCelsius}). Next, \SI{2.5}{\micro\litre} of the rection mixture was transformed into \SI{25}{\micro\litre} Stellar Competent Cells (Takara Bio, 636763). Cells were thawed 15 minutes on ice, plasmid + cell mixture was kept for an additional 30 minutes on ice followed by 45 second heat shot at \SI{42}{\degreeCelsius} and 5 minutes on ice. 
Cells were incubutated for 1 hour at \SI{37}{\degreeCelsius}, single colonies were then grown on Carbencillin plates
over night.
Finally, plasmids were purified using NucleoBond Xtra Maxi Plus Kit (Macherey-Nagel, 740416.50) and eluded at a target
concentration of \SI{4}{\micro\gram\per\micro\litre}.

\paragraph*{Injection control cloning}
Plasmid pCI H2B-RFP
(\href{https://rrid.site/data/record/nif-0000-11872-1/Addgene_92398/resolver}{RRID:Addgene\_92398})\cite{williamsReconstructionGlobalNeural2019}
was ordered from Addgene (\href{https://rrid.site/data/record/nlx_144509-1/SCR_002037/resolver}{RRID:SCR\_002037}),
transformed in Stellar Competent Cells (Takara Bio, 636763). Cells were thawed 15 minutes on ice, plasmid + cell mixture was kept for an additional 30 minutes on ice followed by 45 second heat shot at \SI{42}{\degreeCelsius} and 5 minutes on ice. 
Cells were incubutated for 1 hour at \SI{37}{\degreeCelsius}, single colonies were then grown on Carbencillin plates
over night.
Finally, plasmids were purified using NucleoBond Xtra Maxi Plus Kit (Macherey-Nagel, 740416.50) and eluded at a target
concentration of \SI{4}{\micro\gram\per\micro\litre}.

@@

\subsubsection*{General and specific FOXA2 target regions}
FOXA2 ChIP-seq peaks in HepG2 (\href{https://www.encodeproject.org/files/ENCFF466FCB/}{ENCFF466FCB}) and A549 \href{https://www.encodeproject.org/files/ENCFF686MSH/}{ENCFF686MSH} were downloaded from ENCODE (\href{https://rrid.site/data/record/nlx_144509-1/SCR_006793/resolver}{RRID:SCR\_006793})\cite{encodeprojectconsortiumIntegratedEncyclopediaDNA2012, luoNewDevelopmentsEncyclopedia2020, hitzENCODEUniformAnalysis2023}. 
For each pattern (converted to a PWM) of all neural progenitor pattern cluster motif hits were calculated on the top
1,000 regions of all neural progenitor topics using FIMO\cite{grantFIMOScanningOccurrences2011} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}) using default parameters.
For each sequence the maximum hit, based on the average contribution score of that hit for the topic from which the
region originated, was retained.
Attribution scores were normalized by dividing over the sum of attribution scores over all patterns per sequence.
FOX hits were defined as hits with a value greater than 0.0008 in organoid and 0.0004 in embryo.
To define general hits, the FOX hits were intersected with FOXA2 ChIP-seq peaks in HepG2 and A549 using bedtools (v2.31;
\href{https://rrid.site/data/record/nlx_144509-1/SCR_006646/resolver?q=bedtools&l=bedtools&i=rrid:scr_006646}{RRID:SCR\_006646}) intersect using option -F 0.4 and peaks intersecting with either of the two ChIP-seq peak sets were defined as general.
